taxize - taxonomic search and retrieval in R

### 2013-05-03 inst/doc/taxize_withcode.Rnw
edits to manuscript

Loading into your R session

@@

This is a common task in biology. We often have a list of species names and we want to know if a) we have the most up to date names, b) our names are spelled correctly, and c) if we have common names, we likely need the scientific names. One way to resolve names is via the Global Names Resolver service provided by the Encyclopedia of Life (\url{http://resolver.globalnames.org/}).

@@

# let's take a peek at the data, excluding the data source ID and score columns
temp[,-c(1,4)]
@

Looks like the correct spellings are \emph{Helianthus annuus} and \emph{Homo sapiens}, cool!

Another approach is using the Taxonomic Name Resolution Service via the Taxosaurus API (\url{http://taxosaurus.org/}).

<<tnrs, eval=TRUE>>=
# Lets set our list of species names
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
		"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")

# And we'll call the API with the tnrs function
tnrs(query = mynames)[,-c(5:7)]
@

@@

@@

Another task biologists often face is wanting to get higher taxonomic names for their list of taxa. If you have the higher taxonomy you can put in to context the relationships of your list (i.e., Species A and B are in Family X), as opposed to not knowing that Species A and B are closely related. This also makes it easy to aggregate/standardize data to a specific taxonomic level (eg. family level) or to match data to other databases with different taxonomic resolution (eg. trait-data).

@@

@@

There is an increasingly common use case: many biologists are not adequately trained in reconstructing phylogenies. However, so-called \emph{taxonomic phylogenies} can be constructed from simply knowing the higher taxonomic classification of a set of taxa. There are few taxon groups for which we can get phylogenies simply based on taxonomy; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that lets you input a simple species list, and then taxize does work behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example.

@@

\subsection{What taxa are in children of my taxon of interest?}

>>>>

Loading taxize into your R session

@@

This is a common task in biology. We often have a list of species names and we want to know if a) we have the most up to date names, b) our names are spelled correctly, and c) if we have common names, we likely need the scientific names. One way to resolve names is via the Global Names Resolver (GNR) service provided by the Encyclopedia of Life (\url{http://resolver.globalnames.org/}).

@@

# Let's take a peek at the data, excluding the data source ID and score columns
temp[,-c(1,4)]
@

It looks like the correct spellings are \emph{Helianthus annuus} and \emph{Homo sapiens}. Another approach uses the Taxonomic Name Resolution Service via the Taxosaurus API (\url{http://taxosaurus.org/}).

<<tnrs, eval=TRUE>>=
# A list of species names, some of which are misspelled
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
		"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")

# And we'll call the API with the tnrs function, and remove a few columns
tnrs(query = mynames)[,-c(5:7)]
@

@@

taxize takes the approach that the user should be able to make decisions about what resource to trust, rather than taxize making the decision. Both the EOL GNR and the TNRS services provide data from a variety of data sources. The use may trust a specific data source, thus may want to use the names from that data source.

@@

Another task biologists often face is getting higher taxonomic names for a taxa list. Having the higher taxonomy allows you to put into context the relationships of your species list (i.e., Species A and B are in Family C), as opposed to not knowing that Species A and B are closely related. This also makes it easy to aggregate/standardize data to a specific taxonomic level (e.g., family level) or to match data to other databases with different taxonomic resolution (e.g., trait data).

@@

\subsection{Interactive name selection}

In the previous section we used the function \emph{get_tsn} to get a classification for two plant species.

<<interactive, eval=TRUE>>=
# In this example, this search term has many matches. The function returns a data.frame of the matches, and asks for user input. Enter the row number of the name you want. 
get_tsn(searchterm="Elisabethae", searchtype="sciname")

# You can pass in a long list of taxonomic names
splist <- c("annona cherimola", 'annona muricata', "quercus robur",
  	"shorea robusta", "pandanus patina", "oryza sativa", "durio zibethinus")
get_tsn(searchterm=splist, searchtype="sciname")

# In this example, note that no match at all returns an NA
get_uid(sciname=c("Chironomus riparius", "aaa vva"))
@

<<>>=
text <- '9,562445749
37,66119516
72,9103195
'
scan(dec=",", text=text)
@

@@

Many biologists are not adequately trained in reconstructing phylogenies. However, there are some sources for getting a phylogeny without having to now how to build one; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that lets you input a simple species list, and then taxize works behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example.

@@

Behind the scenes the function \emph{phylomatic\_tree} retrieves a Taxonomic Serial Number (TSN) from ITIS for each species name, then a string is created for each species like this \emph{poaceae/oryza/oryza_sativa} (with format 'family/genus/genus_epithet'). These strings are submitted to the Phylomatic API, and if no errors occur, a phylogeny in newick format is returned. The \emph{phylomatic\_tree} function also cleans up the newick string and converts it to an \textbf{ape} \emph{phylo} object. The output from \emph{phylomatic\_tree} is a \emph{phylo} object, which can be used for plotting, phylogenetic analyses, and more.

\subsection{What taxa are the children of my taxon of interest?}
